---
title: "Lab02"
format: html
editor: visual
---

# 1 Introduction (Loading required packages)

```{r}
#| message: false
#| label: setup-packages

library(tidyverse)
library(gt)
```

# 2 Reconstructing the Table in the ASEAN Statistical Yearbook

## 2.1 Pivot the Data

Asean Meat Production Data:

```{r}
#| message: false
#| label: import-asean-meat-production
#| code-fold: true

asean <- read_csv("asean_meat_production.csv")
gt_preview(asean)
```

In order to pivot the data such that each column represent a country:

```{r}
#| message: false
#| label: pivot-asean-meat-production-via-country
#| code-fold: true

asean <- 
  asean |>
  pivot_wider(names_from = country,values_from = kt)

asean |>
  gt_preview() |>
  tab_options(
    table.font.size = px(9),
    quarto.disable_processing = TRUE
  )
```

## 2.2 Append ASEAN Column

In order to append a new column named "ASEAN"

```{r}
#| message: false
#| label: append-asean-column
#| code-fold: true

asean <-
  asean |>
  rowwise() |>
  mutate(ASEAN = sum(c_across(`Brunei Darussalam`:`Viet Nam`), na.rm = TRUE))

asean |>
  gt_preview() |>
  tab_options(table.font.size = px(9), quarto.disable_processing = TRUE)
```

## 2.3 Insert Livestock and Poultry Rows

Insertion of livestock and poultry rows:

```{r}
#| message: false
#| label: insert-livestock-poultry-rows
#| code-fold: true

asean <-
  asean |>
  pivot_wider(
    names_from = year,
    values_from = `Brunei Darussalam`:ASEAN
  ) |>
  add_row(meat = "Livestock production", .before = 1) |>
  add_row(meat = "Poultry production", .before = 5) |>
  pivot_longer(
    -meat,
    names_to = "country_year",
    values_to = "kt"
  ) |>
  separate_wider_delim(
    country_year,
    delim = "_",
    names = c("country", "year"),
  ) |>
  pivot_wider(names_from = country, values_from = kt)
asean |>
  gt_preview(bottom_n = 9) |>
  tab_options(table.font.size = px(9), quarto.disable_processing = TRUE)
```

## 2.4 Create Basic gt Table

```{r}
#| message: false
#| label: create-basic-gt-table
#| code-fold: true

gt_asean_meat_production <-
  asean |>
  gt(rowname_col = "meat", groupname_col = "year") |>
  tab_options(
    table.font.size = px(9),
    quarto.disable_processing = TRUE
  )

gt_asean_meat_production
```

## 2.5 Add Annotations

Adding annotation to the table

```{r}
#| message: false
#| label: add-annotations-to-gt-table
#| code-fold: true

gt_asean_meat_production <-
  gt_asean_meat_production |>
  tab_header(
    title = "Livestock and Poultry Meat Production in ASEAN, 2018 - 2021",
    subtitle = "(in thousand metric ton)"
  ) |>
  tab_stubhead(
    label = "Animal Type"
  ) |>
  tab_footnote(
    footnote = md("***Note:*** '-' not available at the time of publication")
  ) |>
  tab_source_note(
    source_note = md("***Source:*** FAOSTAT database")
  )

gt_asean_meat_production
```

## 2.6 Format Numbers

```{r}
#| message: false
#| label: format-numbers
#| code-fold: true

gt_asean_meat_production <- fmt_number(
  gt_asean_meat_production,
  columns = `Brunei Darussalam`:ASEAN,
  decimals = 1
)
gt_asean_meat_production
```

## 2.7 Substitute Strings for Missing Data

```{r}
#| message: false
#| label: substitute-strings-for-missing-data
#| code-fold: true

gt_asean_meat_production <-
  gt_asean_meat_production |>
  sub_missing(
    rows = meat %in% c("Livestock production", "Poultry production"),
    missing_text = ""
  ) |>
  sub_missing( # Override the default in "production" rows
    rows = !(meat %in% c("Livestock production", "Poultry production")),
    missing_text = "-"
  )
gt_asean_meat_production
```

## 2.8 Adjust Stub Width

```{r}
#| message: false
#| label: adjust-stub-width
#| code-fold: true

gt_asean_meat_production <-
  gt_asean_meat_production |>
  cols_width(
    stub() ~ px(120)
  )

gt_asean_meat_production
```

## 2.9 Add Summary Rows

For each year, the table in Figure 1.1 displays a summary row containing the total meat production in each column, excluding missing values from the summation. Add these rows to the preliminary table, displaying the number using one decimal and a comma as thousands separator. At this stage, your table should look as follows:

```{r}
#| message: false
#| label: add-summary-rows
#| code-fold: true

gt_asean_meat_production <-
  gt_asean_meat_production |>
  summary_rows(
    fns = "Total meat production" ~ sum(., na.rm = TRUE),
    fmt = ~ fmt_number(., decimals = 1, use_seps = TRUE, sep_mark = ",")
  )

gt_asean_meat_production
```

---

# 3 Styling the Table

## 3.1 Style the Header

```{r}
#| message: false
#| label: style-header
#| code-fold: true

gt_asean_meat_production <-
  gt_asean_meat_production |>
  tab_style(
    style = cell_text(size = px(12), weight = "bold"),
    locations = cells_title(groups="title")
  ) |>
  tab_style(
    style = cell_text(align = "right"),
    locations = cells_title(groups="subtitle")
  )
  

gt_asean_meat_production
```

## 3.2 Style the Stubhead and Column Labels

```{r}
#| message: false
#| label: style-stubhead-column-labels
#| code-fold: true

gt_asean_meat_production <-
  gt_asean_meat_production |>
  tab_style(
    style = list(
      cell_fill(color = "#336636"),
      cell_text(align="center", v_align="middle", color="white", weight="bold"),
      cell_borders(color="white")
    ),
    locations = list(cells_stubhead(), cells_column_labels())
  )

gt_asean_meat_production
```

## 3.3 Style the Row-Group Labels for the Years

```{r}
#| message: false
#| label: style-row-group-labels-for-the-years
#| code-fold: true

gt_asean_meat_production <-
  gt_asean_meat_production |>
  tab_style(
    style = list(
      cell_text(weight="bold", color="white"),
      cell_fill(color="#478846")
    ),
    locations = cells_row_groups()
  )

gt_asean_meat_production
```

## 3.4 Style the Row-Group Labels for Livestock and Poultry Production

```{r}
#| message: false
#| label: style-row-group-labels-for-livestock-poultry-production
#| code-fold: true

gt_asean_meat_production <-
  gt_asean_meat_production |>
  tab_style(
    style = list(
      cell_fill(color = "#bfdeaf"),
      cell_text(weight = "bold"),
      cell_borders(style="hidden")
    ),
    locations = list(
      cells_body(rows = str_ends(meat,"production")),
      cells_stub(rows = str_ends(meat, "production"))
    )
  )


gt_asean_meat_production
```

## 3.5 Style the Summary Labels and Summary Cells

```{r}
#| message: false
#| label: style-summary-labels-summary-cells
#| code-fold: true

gt_asean_meat_production <-
  gt_asean_meat_production |>
  tab_style(
    style = list(
      cell_fill(color = "#478846"),
      cell_text(color = "white", weight = "bold"),
      cell_borders(color = "white")
    ),
    locations = list(
      cells_stub_summary(),
      cells_summary()
    )
  )

gt_asean_meat_production
```

## 3.6 Style All Other Cells
```{r}
#| message: false
#| label: style-all-other-cells
#| code-fold: true

gt_asean_meat_production <-
  gt_asean_meat_production |>
  tab_style(
    style = list(cell_fill(color = "#ebf4ec"), cell_borders(color = "white")),
    locations = list(
      cells_body(
        rows = meat %in% c(
          "Beef and Buffalo Meat",
          "Mutton and Goat Meat",
          "Poultry meat"
        )
      ),
      cells_stub(
        rows = meat %in% c(
          "Beef and Buffalo Meat",
          "Mutton and Goat Meat",
          "Poultry meat"
        )
      )
    )
  ) |>
  tab_style(
    style = list(cell_fill(color = "#d5e9d4"), cell_borders(color = "white")),
    locations = list(
      cells_body(rows = meat == "Pig meat"),
      cells_stub(rows = meat == "Pig meat")
    )
  )
gt_asean_meat_production
```

## 3.7 Reduce Padding
```{r}
#| message: false
#| label: reduce-padding
#| code-fold: true

gt_asean_meat_production <-
  gt_asean_meat_production |>
  tab_options(
    row_group.padding = px(4),
    data_row.padding = px(4),
    summary_row.padding = px(4)
  )
gt_asean_meat_production
```

