---
title: "Lab04"
format: html
editor: visual
theme: lux
---

# 1. Introduction

## 1.1 Load necessary libraries

```{r}
#| label: setup
#| message: false
#| code-fold: true

library(tidyverse)
library(astsa)
library(palmerpenguins)
library(ggbeeswarm)
library(ggmosaic)
```

# 2. Visualizing the Distribution of Quantitative Data: Bill Lengths of Pygoscelis Penguins

## 2.1 Remove Missing Values

Remove the penguins with missing bill length measurements from the `penguins` tibble to avoid ggplot2 warning messages in subsequent tasks.

```{r}
#| label: remove_na
#| message: false
#| code-fold: true

data(penguins)
penguins <- drop_na(penguins, bill_length_mm)

# to ensure bill_length_mm column is clean of missing values
is.na(penguins$bill_length_mm) |> sum() > 0
```

## 2.2 Scatter Plot of Species versus Bill Length

Create a scatter plot displaying bill length on the x-axis and species on the y-axis. For this task and all subsequent ones, customize the axis labels and add a plot title. Additionally, include a plot caption attributing the source as “Gorman et al. (2014)”.

```{r}
#| label: scatter_plot
#| fig-cap: "Basic scatter plot"
#| message: false
#| code-fold: true

ggplot(penguins, aes(x=bill_length_mm, y=species)) +
  geom_point() +
  labs(
    title="Scatter Plot of Species versus Bill Length",
    x="Bill Length (mm)",
    y="Species",
    caption="Source: Gorman et al. (2014)"
  )
```

## 2.3 Why Should You Jitter the Plot?

Which feature of the data suggests that jittering the scatter plot might be beneficial? (Hint: Show that individuals cannot be uniquely identified by their bill length and species.)

Individuals cannot be uniquely identified by their bill length and species, as the following code shows:

```{r}
#| label: jitter_plot_question
#| message: false
#| code-fold: true

anyDuplicated(select(penguins, species, bill_length_mm)) > 0
```

**Note: -** Because of the presence of duplicates, the scatter plot at 2.2 exhibits overplotting; that is, some points occupy exactly the same positions in the plot.

## 2.4 Jitter Plot

Create a jitter plot displaying bill length on the x-axis and species on the y-axis. Reduce the opacity of the points and adjust the `width` and `height` arguments so that points are only jittered in the y-direction while ensuring that the points for different species are clearly separated.

```{r}
#| label: jitter_plot
#| fig-cap: "Jitter plot"
#| message: false
#| code-fold: true

ggplot(penguins, aes(x=bill_length_mm, y=species)) +
  geom_jitter(alpha=0.5, width=0.0, height=0.3) +
  labs(
    title="Jitter Plot of Species versus Bill Length",
    x="Bill Length (mm)",
    y="Species",
    caption="Source: Gorman et al. (2014)"
  )
```

## 2.5 Bee-Swarm Plot

Use the `geom_beeswarm()` function from the ggbeeswarm package to produce a bee-swarm plot illustrating the distribution of bill lengths for each species. Display bill length on the x-axis and species on the y-axis.

```{r}
#| label: bee_swarm_plot
#| fig-cap: "Bee-swarm plot"
#| message: false
#| code-fold: true

ggplot(penguins, aes(x=bill_length_mm, y=species)) +
  geom_beeswarm(size=0.8, cex=2) +
  labs(
    title="Beeswarm Plot of Species versus Bill Length",
    x="Bill Length (mm)",
    y="Species",
    caption="Source: Gorman et al. (2014)"
  )
```

## 2.6 Pros and Cons of Bee-Swarm Plots

> State one advantage and one disadvantage of using a bee-swarm plot rather than a jitter plot for these data?

A bee-swarm plot spreads out the points in clusters of equal or nearly equal data values such that the points are close to each other but do not overlap.

**Advantage:** The bee-swarm layout enables a clearer identification of individual points compared to a jitter plot, where points can overlap due to their random placement. <br>

**Disadvantage:** A potential drawback of bee-swarm plots is that points may need to be smaller or adjacent categories may need to be farther apart to avoid overlapping.

## 2.7 Violin Plot with Boxes

Combine a violin plot and a box plot in one graph. The plot should display bill length on the x-axis and species on the y-axis.

```{r}
#| label: violin_plot_with_boxes
#| fig-cap: "Violin plot with boxes"
#| message: false
#| code-fold: true

ggplot(penguins, aes(x=bill_length_mm, y=species)) +
  geom_boxplot() +
  geom_violin(fill="lightblue", alpha=0.2) +
  labs(
    title="Violin Plot with Box Plot of Species versus Bill Length",
    x="Bill Length (mm)",
    y="Species",
    caption="Source: Gorman et al. (2014)"
  )
```

## 2.8 Frequency Polygons with Rug Plot

Display the bill-length distribution of each species as a frequency polygon, using a bin width of your choice. Combine all three frequency polygons into one coordinate system using a color aesthetic. Place a rug plot below the x-axis to show the distribution of individual data points with corresponding colors.

```{r}
#| label: frequency_polygons_with_rug_plot
#| fig-cap: "Frequency polygon with rug plot"
#| message: false
#| code-fold: true

ggplot(penguins, aes(x=bill_length_mm, color=species)) +
  geom_freqpoly(binwidth = 1) +
  geom_rug( alpha=0.5 ) +
  labs(
    title="Frequency Polygons with Rug Plot of Bill Length",
    x="Bill Length (mm)",
    y="Count",
    color="Species",
    caption="Source: Gorman et al. (2014)"
  )
```

## 2.9 Faceted Frequency Polygons with Rug Plots

Produce a faceted plot that displays the bill-length distribution of each species in a separate facet, using the same bin width as in the previous task. Add a rug plot below each frequency polygon.

```{r}
#| label: faceted_frequency_polygons_with_rug_plots
#| fig-cap: "Faceted histograms with rug plots"
#| message: false
#| code-fold: true

ggplot(penguins, aes(x=bill_length_mm, color=species)) +
  geom_freqpoly(binwidth = 1) +
  geom_rug( alpha = 0.5 ) +
  facet_wrap(~species, ncol=1) +
  labs(
    title="Faceted Frequency Polygons with Rug Plots of Bill Length",
    x="Bill Length (mm)",
    y="Count",
    caption="Source: Gorman et al. (2014)"
  )
```

## 2.10 Pro and Con of Faceting

> State one strength and one weakness of the faceted plot in the previous task compared to the earlier non-faceted frequency polygons with rug plot.

**Advantage:**

Faceting clearly separates the distribution of each category, eliminating the need for color coding. It also reduces the degree of overplotting in the rug plots compared to the non-faceted rug plot

**Disadvantage:**

Faceting can lead to a loss of the overall context of the data, as the distance between different frequency polygons in the plot increases. Thus, the reader needs to shift the view to draw comparisons.

## 2.11 Unmarked Bonus Question: Pairwise *t*-Test

All plots above suggest that the bill length of the Adélie penguin is shorter than that of the gentoo and chinstrap penguins. Let us support this observation quantitatively: 1. Calculate the mean bill length for each species. 2. Perform a pairwise *t*-test using R. Report the *p*-value and interpret the result using a significance level of *a* = 0.05.

1.  Mean bill length by species:

```{r}
#| label: mean_bill_length
#| message: false
#| code-fold: true

penguins_new <- 
  penguins |>
  group_by(species) |>
  mutate(mean = mean(bill_length_mm, na.rm = TRUE)) |>
  distinct(species, mean) 

# show mean bill length of each species
penguins_new
```

***Note:*** As seen in the tibble above, the smallest mean bill length is Adelie's penguin.

2.  Pairwise t-test:

```{r}
#| label: pairwise_t_test
#| message: false
#| code-fold: true

pairwise.t.test(penguins$bill_length_mm, penguins$species)
```

All 𝑝-values are below the usual significance level of 0.05, indicating that the bill lengths of the three species are significantly different from each other. Specifically, Adélie penguins have the shortest bills, while Chinstrap penguins have the longest bills.

------------------------------------------------------------------------

# 3 Visualizing the Distribution of Categorical Data: Travelers on the Titanic

```{r}
#| label: load_titanic
#| message: false
#| code-fold: true

titanic <- read_csv("titanic.csv")
```

## 3.1 Bar Chart of Travelers by Class

Create a bar chart displaying the number of travelers by class (1st, 2nd, 3rd, and crew). Include a plot title and a caption attributing the source as “Encyclopedia Titanica (2023).” Aim for a solution similar to Figure 3.1.

Hint: You may find it easier to aggregate the number of passenger using `count()` before calling `ggplot()`.

```{r}
#| label: bar_chart
#| message: false
#| code-fold: true

titanic_new <-
  titanic |>
  group_by(class) |>
  summarize(count = n())

ggplot(titanic_new, aes(x=count, y=class, label=count)) +
  geom_col(fill=NA, color="black") +
  geom_text(hjust=0, size=3, nudge_x=10) +
  labs(
    title="Titanic Travelers by Class",
    x=NULL,
    y=NULL,
    caption="Source: Encyclopedia Titanica (2023)"
  ) +
  xlim(0, 975) + 
  guides(x="none")
```

## 3.2 Dodged Bar Chart of Travelers by Class and Survival

Use a dodged bar chart to visualize the dependence of the survival rate on the class in which a person traveled. Aim for a solution similar to Figure 3.2.

```{r}
#| labeL: dodged_bar_chart
#| message: false
#| code-fold: true

titanic_new <-
  titanic |>
  group_by(class, survived) |>
  summarize(count = n())

ggplot(titanic_new, aes(x=count, y=class, fill=survived, color=survived, label=count)) +
  geom_col(position="dodge") +
  geom_text(position=position_dodge(width=0.9), vjust=0.5, hjust=-0.03) +
  labs(
    title="Titanic Travelers by Class and Survival",
    x=NULL,
    y=NULL,
    caption="Source: Encyclopedia Titanica (2023)"
  ) +
  xlim(0, 750) +
  guides(x="none")
```

## 3.3 Mosaic Plot

When working with two categorical variables, such as class and survival, mosaic plots provide an alternative to segmented bar charts. Use the `geom_mosaic()` function from the ggmosaic package to create a mosaic plot of the data in `titanic`. Aim for a solution similar to Figure 3.3.

This task requires some hacking skills. Look at the documentation for `geom_mosaic()`. A hint for printing the number of travelers in each category at the center of the corresponding mosaic tile can be found on Stack Overflow.

```{r}
#| label: mosaic_plot
#| message: false
#| warning: false
#| code-fold: true

p <- # Save p as a workaround to determine midpoints in each rectangle
  ggplot(titanic) +
  geom_mosaic(aes(product(class), fill = survived))
p + geom_text(
  data = group_by(ggplot_build(p)$data[[1]], x__class),
  aes(x = (xmin + xmax) / 2, y = (ymin + ymax) / 2, label = .wt)
) +
  labs(
    x = "Class",
    y = "Survived",
    title = "Titanic Travelers by Class and Survival",
    caption = "Source: Encyclopedia Titanica (2023)"
  ) +
  guides(fill = "none")

```

## 3.4 Pros and Cons of Mosaic Plots

> State one advantage and one disadvantage of using the mosaic plot in Figure 3.3 instead of the dodged bar chart in Figure 3.2 for the Titanic data.

**Advantage:**

Unlike in the dodged bar chart, where the lengths of the bars represent absolute numbers, each class occupies the full height of the mosaic plot. Therefore, the mosaic plot provides a clearer visual representation of the proportions of survivors and victims across classes.

**Disadvantage:**

In the dodged bar plot, it is easier to compare the absolute number of survivors across classes because all of the bars are equally thick; hence, the areas are directly proportional to the lengths. The same argument applies, of course, also to the comparison of the absolute number of victims across classes.
