---
title: "Lab04"
format: html
editor: visual
theme: lux
---

# 1. Introduction

## 1.1 Load necessary libraries

```{r}
#| label: setup
#| message: false
#| code-fold: true

library(tidyverse)
library(astsa)
library(palmerpenguins)
library(ggbeeswarm)
library(ggmosaic)
```

# 2. Visualizing the Distribution of Quantitative Data: Bill Lengths of Pygoscelis Penguins

## 2.1 Remove Missing Values
Remove the penguins with missing bill length measurements from the `penguins` tibble to avoid ggplot2 warning messages in subsequent tasks.
```{r}
#| label: remove_na
#| message: false
#| code-fold: true

penguins <- 
  filter(penguins, !is.na(bill_length_mm))

# to ensure bill_length_mm column is clean of missing values
is.na(penguins$bill_length_mm) |> sum() > 0
```

## 2.2 Scatter Plot of Species versus Bill Length
Create a scatter plot displaying bill length on the x-axis and species on the y-axis. For this task and all subsequent ones, customize the axis labels and add a plot title. Additionally, include a plot caption attributing the source as “Gorman et al. (2014)”.
```{r}
#| label: scatter_plot
#| message: false
#| code-fold: true

ggplot(penguins, aes(x=bill_length_mm, y=species)) +
  geom_point() +
  labs(
    title="Scatter Plot of Species versus Bill Length",
    x="Bill Length (mm)",
    caption="Source: Gorman et al. (2014)"
  )
```

## 2.3 Why Should You Jitter the Plot?
Which feature of the data suggests that jittering the scatter plot might be beneficial? (Hint: Show that individuals cannot be uniquely identified by their bill length and species.)
```{r}
#| label: jitter_plot_question
#| message: false
#| code-fold: true

ggplot(penguins, aes(x=bill_length_mm, y=species)) +
  geom_point(position="jitter") +
  labs(
    title="Species versus Bill Length",
    x="Bill Length (mm)",
    caption="Source: Gorman et al. (2014)"
  )
```

**Note: -** In a scatter plot, data are packed together hence with jittering the plot, the data points are more spread out. This makes it easier to see the distribution. X-coord being the bill length while Y-coord being the species

## 2.4 Jitter Plot
Create a jitter plot displaying bill length on the x-axis and species on the y-axis. Reduce the opacity of the points and adjust the `width` and `height` arguments so that points are only jittered in the y-direction while ensuring that the points for different species are clearly separated.
```{r}
#| label: jitter_plot
#| message: false
#| code-fold: true

ggplot(penguins, aes(x=bill_length_mm, y=species)) +
  geom_jitter(alpha=0.5, width=0.6, height=0.3) +
  labs(
    title="Jitter Plot of Species versus Bill Length",
    x="Bill Length (mm)",
    caption="Source: Gorman et al. (2014)"
  )
```

## 2.5 Bee-Swarm Plot
Use the `geom_beeswarm()` function from the ggbeeswarm package to produce a bee-swarm plot illustrating the distribution of bill lengths for each species. Display bill length on the x-axis and species on the y-axis.
```{r}
#| label: bee_swarm_plot
#| message: false
#| code-fold: true

ggplot(penguins, aes(x=bill_length_mm, y=species)) +
  geom_beeswarm() +
  labs(
    title="Beeswarm Plot of Species versus Bill Length",
    x="Bill Length (mm)",
    caption="Source: Gorman et al. (2014)"
  )
```

## 2.6 Pros and Cons of Bee-Swarm Plots
> State one advantage and one disadvantage of using a bee-swarm plot rather than a jitter plot for these data?

**Advantage: -** Bee-Swarm plots are useful for visualizing the distribution of a quantitative variable across different categories. They are especially useful when the number of data points is large.<br> **Disadvantage: -** Bee-Swarm plots can be computationally expensive to create and may not be suitable for very large datasets. With large datasets, meaning there will be a lot of data points, the data points being like a bee will be overlapped and hence the plot will be hard to interpret.

## 2.7 Violin Plot with Boxes
Combine a violin plot and a box plot in one graph. The plot should display bill length on the x-axis and species on the y-axis.
```{r}
#| label: violin_plot_with_boxes
#| message: false
#| code-fold: true

ggplot(penguins, aes(x=bill_length_mm, y=species)) +
  geom_boxplot() +
  geom_violin(fill="lightblue", alpha=0.2) +
  labs(
    title="Violin Plot with Box Plot of Species versus Bill Length",
    x="Bill Length (mm)",
    caption="Source: Gorman et al. (2014)"
  )
```

## 2.8 Frequency Polygons with Rug Plot
Display the bill-length distribution of each species as a frequency polygon, using a bin width of your choice. Combine all three frequency polygons into one coordinate system using a color aesthetic. Place a rug plot below the x-axis to show the distribution of individual data points with corresponding colors.
```{r}
#| label: frequency_polygons_with_rug_plot
#| message: false
#| code-fold: true

ggplot(penguins, aes(x=bill_length_mm, color=species)) +
  geom_freqpoly(bins=20) +
  geom_rug(aes(y=0), position=position_jitter(height=0) ) +
  labs(
    title="Frequency Polygons with Rug Plot of Bill Length",
    x="Bill Length (mm)",
    y="Frequency",
    caption="Source: Gorman et al. (2014)"
  )
```

## 2.9 Faceted Frequency Polygons with Rug Plots
Produce a faceted plot that displays the bill-length distribution of each species in a separate facet, using the same bin width as in the previous task. Add a rug plot below each frequency polygon.
```{r}
#| label: faceted_frequency_polygons_with_rug_plots
#| message: false
#| code-fold: true

ggplot(penguins, aes(x=bill_length_mm, color=species)) +
  geom_freqpoly(bins=20) +
  geom_rug(aes(y=0), position=position_jitter(height=0) ) +
  facet_wrap(~species, ncol=1) +
  labs(
    title="Faceted Frequency Polygons with Rug Plots of Bill Length",
    x="Bill Length (mm)",
    y="Frequency",
    caption="Source: Gorman et al. (2014)"
  )
```

## 2.10 Pro and Con of Faceting
> State one strength and one weakness of the faceted plot in the previous task compared to the earlier non-faceted frequency polygons with rug plot.

**Advantage:** - Faceting allows you to compare the distribution of a quantitative variable across different categories. This can be useful for identifying patterns or trends that may not be apparent when the data are combined.<br> **Disadvantage:** - Faceting is not recommended for time series data when doing a comparison of how the trendline moves over time. Combining each line into one coordinate system will make it easier to see the comparison.

## 2.11 Unmarked Bonus Question: Pairwise *t*-Test
All plots above suggest that the bill length of the Adélie penguin is shorter than that of the gentoo and chinstrap penguins. Let us support this observation quantitatively:
1. Calculate the mean bill length for each species.
2. Perform a pairwise *t*-test using R. Report the *p*-value and interpret the result using a significance level of *a* = 0.05.

```{r}
#| label: mean_bill_length
#| message: false
#| code-fold: true

penguins_new <- 
  penguins |>
  group_by(species) |>
  mutate(mean_bill_length = mean(bill_length_mm, na.rm = TRUE)) |>
  distinct(species, mean_bill_length) 

# show mean bill length of each species
penguins_new
```

***Note:*** As seen in the tibble above, the smallest mean bill length is Adelie's penguin.

```{r}
#| label: pairwise_t_test
#| message: false
#| code-fold: true

# Clean the data
penguins_clean <- penguins |>
  filter(!is.na(bill_length_mm) & !is.na(species))

# Perform pairwise t-test
pairwise_results <- pairwise.t.test(
  penguins_clean$bill_length_mm,
  penguins_clean$species,
  p.adjust.method = "none"
)

p_values <- pairwise_results$p.value
p_values
alpha <- 0.05
significant_comparisons <- p_values < alpha
significant_comparisons
```

------------------------------------------------------------------------

# 3 Visualizing the Distribution of Categorical Data: Travelers on the Titanic

```{r}
#| label: load_titanic
#| message: false
#| code-fold: true

titanic <- read_csv("titanic.csv")
```

## 3.1 Bar Chart of Travelers by Class
Create a bar chart displaying the number of travelers by class (1st, 2nd, 3rd, and crew). Include a plot title and a caption attributing the source as “Encyclopedia Titanica (2023).” Aim for a solution similar to Figure 3.1.

Hint: You may find it easier to aggregate the number of passenger using `count()` before calling `ggplot()`.
```{r}
#| label: bar_chart
#| message: false
#| code-fold: true

titanic_new <-
  titanic |>
  group_by(class) |>
  summarize(count = n())

ggplot(titanic_new, aes(x=count, y=class)) +
  geom_col(fill=NA, color="black", just=0.5) +
  geom_text(aes(label=count), hjust=-0.2) +
  labs(
    title="Titanic Travelers by Class",
    x="Traveler Count",
    caption="Source: Encyclopedia Titanica (2023)"
  )
```

## 3.2 Dodged Bar Chart of Travelers by Class and Survival
Use a dodged bar chart to visualize the dependence of the survival rate on the class in which a person traveled. Aim for a solution similar to Figure 3.2.
```{r}
#| labeL: dodged_bar_chart
#| message: false
#| code-fold: true

titanic_new <-
  titanic |>
  group_by(class, survived) |>
  summarize(count = n())

ggplot(titanic_new, aes(x=count, y=class, fill=survived)) +
  geom_col(position="dodge") +
  geom_text(aes(label=count, color=survived), position=position_dodge(width=0.9), vjust=0.5, hjust=-0.03) +
  labs(
    title="Titanic Travelers by Class and Survival",
    x="Traveler Count",
    caption="Source: Encyclopedia Titanica (2023)"
  )
```

## 3.3 Mosaic Plot
When working with two categorical variables, such as class and survival, mosaic plots provide an alternative to segmented bar charts. Use the `geom_mosaic()` function from the ggmosaic package to create a mosaic plot of the data in `titanic`. Aim for a solution similar to Figure 3.3.

This task requires some hacking skills. Look at the documentation for `geom_mosaic()`. A hint for printing the number of travelers in each category at the center of the corresponding mosaic tile can be found on Stack Overflow.
```{r}
#| label: mosaic_plot
#| message: false
#| warning: false
#| code-fold: true

p <- ggplot(titanic) +
  geom_mosaic(aes(x = product(class), fill = survived)) +
  geom_mosaic_text(aes(x = ""))

p + geom_text(data = ggplot_build(p)$data[[1]],
              aes(x = (xmin+xmax)/2, y = (ymin+ymax)/2, label = .wt)) +
  labs(title = "Titanic Travelers by Class and Survival",
        x = "Class",
        y = "Survived",
       caption = "Source: Encyclopedia Titanica (2023)") +
  scale_fill_discrete(guide = "none")

```

## 3.4 Pros and Cons of Mosaic Plots
> State one advantage and one disadvantage of using the mosaic plot in Figure 3.3 instead of the dodged bar chart in Figure 3.2 for the Titanic data.

**Advantage:** - The mosaic plot provides a clearer view of how survival rates vary across passenger classes, making comparisons easier.<br> **Disadvantage:** - It might not give as precise numerical information as a dodged bar chart.
