---
title: "lecture_2"
format: html
editor: visual
---

# Required Packages

```{r}
#| message: false
#| label: setup
library(tidyverse)
library(gt)
```

# Pivoting

New York City mayoral election example:

```{r}
#| message: false
#| label: import-ny-mayor

ny_mayor <- read_csv("ny_mayor_wide.csv")
ny_mayor
```

This data frame has `r nrow(ny_mayor)` rows and `r ncol(ny_mayor)` columns which is relatively wide. We can pivot this data frame to make it longer.

```{r}
#| label: pivot-longer

ny_mayor_long <-
  pivot_longer(ny_mayor,
    cols = c("de_blasio", "malliotakis", "other"),
    names_to = "candidate",
    values_to = "votes"
  )

ny_mayor_long
```

We can also pivot data frames into wider format:

```{r}
#| label: pivot-wider

ny_mayor_wider <- pivot_wider(
  ny_mayor_long,
  names_from = candidate,
  values_from = votes
)

ny_mayor_wider
```

# Pivoting the ASEAN data

```{r}
#| label: import-asean

trade <- read_csv("asean_trade_by_country_and_mode_of_transport.csv")
trade
```

`trade` is currently in a wide format with `r nrow(trade)` rows and `r ncol(trade)` columns. For easier readability, we want to make the data wider.

```{r}
#| label: pivot-wider-asean

trade_wide <- pivot_wider(
  trade,
  names_from = mode,
  values_from = million_usd
)
trade_wide
```

# Create Basic gt Table

```{r}
gt_trade <- 
  trade_wide |>
  gt(rowname_col = "country", groupname_col = "year") |>
  tab_options(
    table.font.size = px(10),
    quarto.disable_processing = TRUE
  )

gt_trade
```

# Add Annotations

```{r}
gt_trade <-
  gt_trade |>
  tab_header(
    title = "ASEAN Trade by Country and Mode of Transport",
    subtitle = "Values in million USD"
  ) |>
  tab_spanner(
    label = "Mode",
    columns = Air:Unknown
  ) |>
  tab_stubhead(
    label = "Country"
  ) |>
  tab_footnote(
    footnote = "Data available in total",
    locations = cells_stub(c("Singapore", "Viet Nam"))
  ) |>
  tab_source_note(source_note = "ASEAN Secretariat (2024)")

gt_trade
```

# Format data

```{r}
#| label: format-data

gt_trade <-
  gt_trade |>
  fmt_number(decimals = 1) |>
  sub_missing()

gt_trade
```

# Modify columns

```{r}
#| label: modify-columns

gt_trade <- cols_width(
  gt_trade,
  "Not elsewhere classified" ~ px(90),
)

gt_trade
```

# Add summary rows

```{r}
#| label: add-summary-rows

gt_trade <-
  gt_trade |>
  summary_rows(
    fns = "Annual total" ~ sum(., na.rm = TRUE),
    fmt = ~ fmt_number(., decimals = 1)
  )

gt_trade
```

# Styling

```{r}
#| label: styling
# color url: http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf

gt_trade <-
  gt_trade |>
  tab_style(
    style = list(
      cell_fill(color = "gray90"),
      cell_text(weight = "bold", style = "italic")
    ),
    locations = cells_title()
  ) |>
  tab_style(
    style = cell_fill(color = "aliceblue"),
    locations = list(cells_stubhead(), cells_stub())
  )

gt_trade
```
